Exploring Anomalies in Pacific Northwest Special Education Trends

### 2020-12-07 scripts/Final_project.Rmd
Please please please work
edited out script, edited citation, changed references

When policy is enacted, the intended outcomes are often not realized for months, years, or decades [@gu-2018]. For example, changing school district policy on professional development to align with research-informed practices may be constructed to increase student academic achievement; though, there is insufficient evidence pointing to how long professionalized intervention with teachers will take to increase student academic scores [@luke_2009]. Longitudinal data helps to understand the impact of policy change because of the year-over-year comparisons that illustrate differences, inflection points, or anomalies of interest [@michelmore_2017].

@@

cleanest_primo2 <- cleanest_primo [!(cleanest_primo$disability=="TOTAL (SUM OF ALL OF THE ABOVE)" |cleanest_primo$disability=="OTHER HEALTH IMPAIRMENT" | cleanest_primo$disability=="SPECIFIC LEARNING DISABILITY" | cleanest_primo$disability=="SPECIFIC LEARNING DISABILITY"),]
cleanest_primo2

@@

fresh_primo %>% 
  group_by(year, disability) %>% 
  summarize(total=sum(frequency, na.rm=TRUE))

@@

There are two major outcomes of interest observed with the data sets when disaggregated by rural and urban school districts. First, the trends of rural student exceptionality designations followed urban trends closely. The vast majority of urban school districts did not differ in their post-2016 prevalence rates for exceptionality designations (ps > 0.05). However, for diagnostically malleable categories (e.g., Moderate Behavioural designation) in rural school districts, 2016 was an inflection point for an increase in prevalence rates. When comparing prevalence rates between 2013-2016 (pre-2016 group) and 2017-2019 (post-2016 group) cohorts of rural students, the prevalence rate increased for the following student designation categories: High Behavioural Interventions (t(239) = -2.113, p = 0.036, g = 0.261, CI[-0.504, -0.019]), Learning Disability (t(262) = -1.984, p = 0.483, g = 0.230, CI[-0.453, -0.007]), and Chronic Health (t(255) = -2.068, p = 0.040, g = 0.247, CI[-0.479, -0.015]). Note that the exceptionality category Mild Behavioural Intervention did not result with the same small, detectable effect size (Hedges’ g > 0.2, and g < 0.5), but did show an increasing trend post-2016 (t(195) = -0.879, p = 0.380, g = 0.120, CI[-0.383, 0.144]).

>>>>

When policy is enacted, the intended outcomes are often not realized for months, years, or decades [@gu_2018]. For example, changing school district policy on professional development to align with research-informed practices may be constructed to increase student academic achievement; though, there is insufficient evidence pointing to how long professionalized intervention with teachers will take to increase student academic scores [@luke_2009]. Longitudinal data helps to understand the impact of policy change because of the year-over-year comparisons that illustrate differences, inflection points, or anomalies of interest [@michelmore_2017].

@@

cleanest_primo2 <- cleanest_primo [!(cleanest_primo$disability=="TOTAL (SUM OF ALL OF THE ABOVE)" |cleanest_primo$disability=="OTHER HEALTH IMPAIRMENT" | cleanest_primo$disability=="SPECIFIC LEARNING DISABILITY" | cleanest_primo$disability=="SPECIFIC LEARNING DISABILITY"),]
#cleanest_primo2

@@

#fresh_primo %>% 
  #group_by(year, disability) %>% 
  #summarize(total=sum(frequency, na.rm=TRUE))

@@

There are two major outcomes of interest observed with the data sets when disaggregated by rural and urban school districts. First, the trends of rural student exceptionality designations followed urban trends closely. The vast majority of urban school districts did not differ in their post-2016 prevalence rates for exceptionality designations (*p*s > 0.05). However, for diagnostically malleable categories (e.g., Moderate Behavioural designation) in rural school districts, 2016 was an inflection point for an increase in prevalence rates. When comparing prevalence rates between 2013-2016 (pre-2016 group) and 2017-2019 (post-2016 group) cohorts of rural students, the prevalence rate increased for the following student designation categories: High Behavioural Interventions (*t*(239) = -2.113, *p* = 0.036, *g* = 0.261, CI[-0.504, -0.019]), Learning Disability (*t*(262) = -1.984, *p* = 0.483, *g* = 0.230, CI[-0.453, -0.007]), and Chronic Health (*t*(255) = -2.068, *p* = 0.040, *g* = 0.247, CI[-0.479, -0.015]). Note that the exceptionality category Mild Behavioural Intervention did not result with the same small, detectable effect size (Hedges’ *g* > 0.2, and *g* < 0.5), but did show an increasing trend post-2016 (*t*(195) = -0.879, *p* = 0.380, *g* = 0.120, CI[-0.383, 0.144]).
